Packaging research artefacts with RO-Crate

### 2021-12-28 latex/ro-crate.tex
Updates from Overleaf
references
moved future work to new section
include future work
future work as part of conclusion
corrected ordering of affiliations
Typeset in LaTeX2e by VTeX Ltd., Vilnius, Lithuania

%% ro-crate paper
%% (c) 2020-2021 The University of Manchester
%% 
%% SPDX-License-Identifier: CC-BY-4.0

% add. options: [seceqn,secthm,crcready]
\documentclass[ds]{iosart2x}

%%%%%%%%%%% Put your definitions here

% From https://www.overleaf.com/latex/templates/using-the-stix2-opentype-fonts-with-lualatex-or-xelatex/hdfvhzqmhnpx
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{hyperref}
\usepackage{fontspec}
\usepackage{unicode-math}
% \usepackage{graphicx}

\DeclareGraphicsExtensions{.pdf,.png,.svg}
\graphicspath{../content/images/}
\usepackage[citations,definitionLists,fencedCode,footnotes,hybrid,pipeTables,smartEllipses,tableCaptions,underscores]{markdown}
\def\markdownOptionOutputDir{../output}

% for table resizing
\usepackage{booktabs}
\usepackage{multirow}

%%%%%%%%%%% End of definitions

\pubyear{2021}
\volume{0}
\firstpage{1}
\lastpage{1}

@@

% Welcome to the 2000s with UNICODE fonts!
% Adapted from
% https://www.overleaf.com/learn/latex/Articles%2FOpenType-based_math_typesetting:_An_introduction_to_the_STIX2_OpenType_fonts
\setmathfont{STIX2Math}[
    Extension={.otf},
    Path=./STIX2fonts/,
    Scale=1
]
\setmainfont{STIX2Text}[
    Extension={.otf},
    Path=./STIX2fonts/,
    UprightFont={*-Regular},
    BoldFont={*-Bold},
    ItalicFont={*-Italic},
    BoldItalicFont={*-BoldItalic}
]

% https://typeof.net/Iosevka/ mono font has 
% all; the logic symbols we need
\setmonofont{iosevka-slab}[
    Extension={.ttf},
    Path=./ttf-iosevka-slab-6.1.3/,
    Scale=MatchLowercase,
    UprightFont={*-light},
    ItalicFont={*-lightitalic}
]

@@

%\pretitle{}
\title{Packaging research artefacts with RO-Crate}
\runtitle{Packaging research artefacts with RO-Crate}
%\subtitle{}

% For one author:
%\author{\inits{N.}\fnms{Name1} \snm{Surname1}\ead[label=e1]{first@somewhere.com}\ead[label=o1,orcid]{0000-0000-0000-0001}}
%\address{Department first, \orgname{University or Company name},
%Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1,o1}}

% Two or more authors:
\author[SSR1,SSR2]{\inits{S.}\fnms{Stian}
    \snm{Soiland-Reyes}\ead[label=eSSR]{soiland-reyes@manchester.ac.uk}\ead[label=oSSR,orcid]{0000-0001-9842-9718}%
    \thanks{Corresponding author. \printead{eSSR}.}}, 
\author[PS]{\inits{P.}\fnms{Peter}
    \snm{Sefton}\ead[label=ePS]{Peter.Sefton@uts.edu.au}\ead[label=oPS,orcid]{0000-0002-3545-944X}},
\author[EOC]{\inits{E.}\fnms{Eoghan}
    \snm{Ó Carragáin}\ead[label=eEOC]{eoghan.ocarragain@ucc.ie}\ead[label=oEOC,orcid]{0000-0001-8131-2150}},
\author[MC]{\inits{M.}\fnms{Mercè}
    \snm{Crosas}\ead[label=eMC]{mcrosas@g.harvard.edu}\ead[label=oMC,orcid]{0000-0003-1304-1939}},
\author[LJC]{\inits{LJ.}\fnms{Leyla Jael}
    \snm{Castro}\ead[label=eLJC]{ljgarcia@zbmed.de}\ead[label=oLJC,orcid]{0000-0003-3986-0510}},
\author[FC]{\inits{F.}\fnms{Frederik}
    \snm{Coppens}\ead[label=eFC]{frcop@psb.vib-ugent.be}\ead[label=oFC,orcid]{0000-0001-6565-5145}},
\author[JMF]{\inits{JM.}\fnms{José M.}
    \snm{Fernández}\ead[label=eJMF]{jose.m.fernandez@bsc.es}\ead[label=oJMF,orcid]{0000-0002-4806-5140}},
\author[DG]{\inits{D.}\fnms{Daniel}
    \snm{Garijo}\ead[label=eDG]{dgarijo@fi.upm.es}\ead[label=oDG,orcid]{0000-0003-0454-7145}},
\author[MLR]{\inits{M.}\fnms{Marco}
    \snm{La Rosa}\ead[label=eMLR]{m@lr.id.au}},
\author[SL]{\inits{S.}\fnms{Simone}
    \snm{Leo}\ead[label=eSL]{simone.leo@crs4.it}\ead[label=oSL,orcid]{0000-0001-8271-5429}},
\author[MP]{\inits{M.}\fnms{Marc}
    \snm{Portier}\ead[label=eMP]{marc.portier@vliz.be}\ead[label=oMP,orcid]{0000-0002-9648-6484}},
\author[AT]{\inits{A.}\fnms{Ana}
    \snm{Trisovic}\ead[label=eMP]{ana.trisovic@gmail.com}\ead[label=oAT,orcid]{0000-0003-1991-0533}},
\author[RO]{RO-Crate Community},
\author[PG]{\inits{P.}\fnms{Paul}
    \snm{Groth}\ead[label=ePG]{p.t.groth@uva.nl}\ead[label=oPG,orcid]{0000-0003-0183-6910}},
\author[CG]{\inits{C.}\fnms{Carole}
    \snm{Goble}\ead[label=eCG]{carole.goble@manchester.ac.uk}\ead[label=oCG,orcid]{0000-0003-1219-2137}}

\address[SSR1]{Computer Science, 
    \orgname{The University of Manchester}, \cny{UK}}
\address[SSR2]{Informatics Institute, \orgname{University of Amsterdam},
    \cny{The Netherlands}\printead[presep={\\}]{eSSR,oSSR}}
\address[PS]{Faculty of Science, \orgname{University Technology Sydney},
    \cny{Australia}\printead[presep={\\}]{ePS,oPS}}
\address[EOC]{\orgname{University College Cork},
    \cny{Ireland}\printead[presep={\\}]{oEOC}}
\address[MC]{Institute for Quantitative Social Science, 
    \orgname{Harvard University}, Cambridge, MA,  \cny{USA}\printead[presep={\\}]{oMC}}
\address[LJC]{\orgname{ZB MED Information Centre for Life Sciences}, 
    Cologne, \cny{Germany}\printead[presep={\\}]{oLJC}}
\address[FC]{\orgname{VIB-UGent Center for Plant Systems Biology}, 
    Gent, \cny{Belgium}\printead[presep={\\}]{oFC}}
\address[JMF]{\orgname{Barcelona Supercomputing Center}, 
    Barcelona, \cny{Spain}\printead[presep={\\}]{oJMF}}
\address[DG]{Ontology Engineering Group, 
    \orgname{Universidad Politécnica de Madrid}, Madrid, \cny{Spain}\printead[presep={\\}]{oDG}}
\address[MLR]{\printead{eMLR}}
\address[SL]{\orgname{Center for Advanced Studies, Research, and Development in Sardinia} (CRS4), Pula (CA),
    \cny{Italy}\printead[presep={\\}]{oSL}}
\address[MP]{\orgname{Vlaams Instituut voor de Zee}, Oostende,
    \cny{Belgium}\printead[presep={\\}]{oMP}}
\address[AT]{Institute for Quantitative Social Science,
    \orgname{Harvard University}, Cambridge, MA, \cny{USA}\printead[presep={\\}]{oAT}}
\address[RO]{\url{https://www.researchobject.org/ro-crate/community} (see appendix B)}
\address[PG]{Informatics Institute, \orgname{University of Amsterdam},
    \cny{The Netherlands}\printead[presep={\\}]{ePG,oPG}}
\address[CG]{Computer Science, \orgname{The University of Manchester},
    \cny{UK}\printead[presep={\\}]{eCG,oCG}}

%\begin{review}{editor}
%\reviewer{\fnms{First} \snm{Editor}\ead[label=oe1,orcid]{}\printead[presep={\ }]{oe1}}
%\reviewer{\fnms{Second} \snm{Editor}\ead[label=oe2,orcid]{}\printead[presep={\ }]{oe2}}
%\end{review}
%\begin{review}{solicited}
%\reviewer{\fnms{First} \snm{Solicited reviewer}\ead[label=os1,orcid]{}\printead[presep={\ }]{os1}}
%\end{review}

\newpage % Because the author affiliations are too long!
\begin{abstract}
\markdownInput{../content/01.abstract.md}
\end{abstract}

\begin{keyword}
\kwd{Data publishing}
\kwd{Data packaging}
\kwd{FAIR}
\kwd{Linked Data}
\kwd{Metadata}
\kwd{Reproducibility}
\kwd{Research Object}
\end{keyword}

\end{frontmatter}

%%%%%%%%%%% The article body starts:

% Every *.md file in content/ in order, except 00, 90
\markdownInput{../content/10.background.md}

\markdownInput{../content/20.rocrate.md}
\markdownInput{../content/21.conceptual.md}
\markdownInput{../content/22.implementation.md}
\markdownInput{../content/23.community.md}

\input{../content/31.tooling-table.tex}
%\markdownInput{../content/31.tooling-table.md}
\markdownInput{../content/30.tooling.md}

\markdownInput{../content/40.inuse.md}
\markdownInput{../content/41.workflows.md}
\markdownInput{../content/42.regulatory.md}
\markdownInput{../content/43.culturalheritage.md}
\markdownInput{../content/44.dmp.md}

\markdownInput{../content/50.relatedwork.md}
\markdownInput{../content/60.conclusion.md}
\markdownInput{../content/70.acknowledgements.md}

%%%%%%%%%%% The bibliography starts:
\nocite{*} 
\bibliographystyle{ios1}
\bibliography{ro-crate}

%%%% Appendix

\appendix
\markdownInput{../content/80.formal-definition.md}
% TODO: Recode formal definition to LaTeX math mode?
%\include{../content/80.formal-definition}

\markdownInput{../content/89.community.md}

\end{document}

>>>>

%BeginFileInfo
%%Publisher=IOS
%%Project=DS
%%Manuscript=DS210053
%%MS position=
%%Stage=201
%%TID=imikolaityte
%%Pages=0
%%Format=2019
%%Distribution=vtex
%%Destination=DVI
%%PDF type=
%%DVI.Maker=luadvi
%%PS.Maker=luaps
%%History1=2021.12.28 09:11
%EndFileInfo
%
% Journal: Data Science (DS), IOS Press
% Typeset in LaTeX2e by VTeX Ltd., Vilnius, Lithuania
%
%Spelling_date
%\RequirePackage{float} % when algorithm.sty used in manuscript
\documentclass[ds,v1.1.2,openaccess]{iosart2x}%% ,seceqn
\usepackage{mathbh}
%
%\usepackage{}
\usepackage{tmultirow}
\usepackage{listings}
%
\PROOF
%\PROOF
%\CRC

\volume{-1}
\issue{-1}
\pubyear{2021}
\firstpage{1}
\lastpage{42}
\articletype{research-article} %%
%http://www.vtex.vmt/docs/wiki/index.php/IOS_Press_instrukcijos_%28Kaunas%29#Straipsnio_tipas_.28.5Carticletype.7B.7D.29
\aid{DS210053}% Updated by VTEXPTS2LaTeX.exe, 28.12.2021 09:11
\doi{10.3233/DS-210053}% Updated by VTEXPTS2LaTeX.exe, 28.12.2021 09:11
\vtexed{Ingrida}

\startlocaldefs
\setpkgattr{license}{type}{CC@BY} %% CC@BY@NC
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% indentation after theorem-like environments - auto
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\newtheorem*{}{} %% unnumbered
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\theoremstyle{plain}
%%\newtheorem{}{}%[]
%%\theoremstyle{definition}
%%\newtheorem{}{}%[]
\endlocaldefs
%

@@

%

@@

%
\pretitle[XML]{Research Article}
%\pretitle{}
\title{Packaging research artefacts with RO-Crate}
%\title{\thanks{}}
\runtitle{Packaging research artefacts with RO-Crate}
%\subtitle{}
%
%%% initials without spaces
%
% {\inits{}\fnms{} \snm{} \suffix{}\ead[label=e?]{}
%\ead[label=p1,phone]{}\ead[label=f1,fax]{}\ead[label=o?,orcid]{}
%\thanks{}}
%
\begin{aug}
\author[A,B]{\inits{S.}\fnms{Stian} \snm{Soiland-Reyes}\ead
[label=e1]{soiland-reyes@manchester.ac.uk}\ead
[label=o1,orcid]{0000-0001-9842-9718}\thanks{Corresponding author.
\printead{e1}.}}
\author[C]{\inits{P.}\fnms{Peter} \snm{Sefton}\ead
[label=e2]{Peter.Sefton@uts.edu.au}\ead[label=o2,orcid]{0000-0002-3545-944X}}
\author[D]{\inits{M.}\fnms{Merc\`{e}} \snm{Crosas}\ead
[label=o3,orcid]{0000-0003-1304-1939}}
\author[E]{\inits{L.J.}\fnms{Leyla Jael} \snm{Castro}\ead
[label=o4,orcid]{0000-0003-3986-0510}}
\author[F]{\inits{F.}\fnms{Frederik} \snm{Coppens}\ead
[label=o5,orcid]{0000-0001-6565-5145}}
\author[G]{\inits{J.M.}\fnms{Jos\'{e} M.} \snm{Fern\'{a}ndez}\ead
[label=o6,orcid]{0000-0002-4806-5140}}
\author[H]{\inits{D.}\fnms{Daniel} \snm{Garijo}\ead
[label=o7,orcid]{0000-0003-0454-7145}}
\author[I]{\inits{B.}\fnms{Bj\"{o}rn} \snm{Gr\"{u}ning}\ead
[label=o8,orcid]{0000-0002-3079-6586}}
\author[J]{\inits{M.}\fnms{Marco} \snm{La Rosa}\ead
[label=o9,orcid]{0000-0001-5383-6993}}
\author[K]{\inits{S.}\fnms{Simone} \snm{Leo}\ead
[label=o10,orcid]{0000-0001-8271-5429}}
\author[L]{\inits{E.}\fnms{Eoghan} \snm{\'{O} Carrag\'{a}in}\ead
[label=o11,orcid]{0000-0001-8131-2150}}
\author[M]{\inits{M.}\fnms{Marc} \snm{Portier}\ead
[label=o12,orcid]{0000-0002-9648-6484}}
\author[N]{\inits{A.}\fnms{Ana} \snm{Trisovic}\ead
[label=o13,orcid]{0000-0003-1991-0533}}
\author[O]{\snm{RO-Crate Community}\ead
[label=u1,url]{https://www.researchobject.org/ro-crate/community}}
\author[P]{\inits{P.}\fnms{Paul} \snm{Groth}\ead
[label=e15]{p.t.groth@uva.nl}\ead[label=o15,orcid]{0000-0003-0183-6910}}
\author[Q]{\inits{C.}\fnms{Carole} \snm{Goble}\ead
[label=e16]{carole.goble@manchester.ac.uk}\ead
[label=o16,orcid]{0000-0003-1219-2137}}
%
%%%%%% printead order: TEL FAX E-MAIL URL
%\thanks{Corresponding author: name surname, address. \printead{}.}
%\thanks{Corresponding author. \printead{}.}
%\thanks{}%% Move support info to acks
%
\runauthor{S. Soiland-Reyes et al.} %% auto
% {\orgname{} \cny{}\printead[presep={\\}]{}}
\address[A]{Department of Computer Science, \orgname{The University of
Manchester}, \cny{UK}\printead[presep={\\}]{e1,o1}}
\address[B]{Informatics Institute, \orgname{University of Amsterdam},
\cny{The Netherlands}}
\address[C]{Faculty of Science, \orgname{University Technology Sydney},
\cny{Australia}\printead[presep={\\}]{e2,o2}}
\address[D]{Institute for Quantitative Social Science, \orgname{Harvard
University}, Cambridge, MA, \cny{USA}\printead[presep={\\}]{o3}}
\address[E]{\orgname{ZB MED Information Centre for Life Sciences},
Cologne, \cny{Germany}\printead[presep={\\}]{o4}}
\address[F]{\orgname{VIB-UGent Center for Plant Systems Biology}, Gent,
\cny{Belgium}\printead[presep={\\}]{o5}}
\address[G]{\orgname{Barcelona Supercomputing Center}, Barcelona, \cny
{Spain}\printead[presep={\\}]{o6}}
\address[H]{Ontology Engineering Group, \orgname{Universidad Polit\'
{e}cnica de Madrid}, Madrid, \cny{Spain}\printead[presep={\\}]{o7}}
\address[I]{Bioinformatics Group, Department of Computer Science,
\orgname{Albert-Ludwigs-University Freiburg}, Freiburg, \cny
{Germany}\printead[presep={\\}]{o8}}
\address[J]{\orgname{PARADISEC}, Melbourne, \cny{Australia}\printead
[presep={\\}]{o9}}
\address[K]{\orgname{Center for Advanced Studies, Research, and
Development in Sardinia} (CRS4), Pula (CA), \cny{Italy}\printead
[presep={\\}]{o10}}
\address[L]{\orgname{University College Cork}, \cny{Ireland}\printead
[presep={\\}]{o11}}
\address[M]{\orgname{Vlaams Instituut voor de Zee}, Oostende, \cny
{Belgium}\printead[presep={\\}]{o12}}
\address[N]{Institute for Quantitative Social Science, \orgname{Harvard
University}, Cambridge, MA, \cny{USA}\printead[presep={\\}]{o13}}
\address[O]{\printead{u1}~(see Appendix B)}
\address[P]{Informatics Institute, \orgname{University of Amsterdam},
\cny{The Netherlands}\printead[presep={\\}]{e15,o15}}
\address[Q]{Department of Computer Science, \orgname{The University of
Manchester}, \cny{UK}\printead[presep={\\}]{e16,o16}}
\end{aug}
%
%%%%%%
%
\begin{review}{editor}
\reviewer{\fnms{Silvio} \snm{Peroni}\ead
[label=oe1,orcid]{0000-0003-0530-4305}\printead[presep={\ }]{oe1}}
%\reviewer{\fnms{} \snm{}\ead[label=oe2,orcid]{}\printead[presep={\
%}]{oe2}}
\end{review}

\begin{review}{solicited}
\reviewer{\fnms{Angelo A.} \snm{Salatino}\ead
[label=os1,orcid]{0000-0002-4763-3943}\printead[presep={\ }]{os1}}
\reviewer{\fnms{Alessia} \snm{Bardi}\ead
[label=os2,orcid]{0000-0002-1112-1292}\printead[presep={\ }]{os2}}
\reviewer{\snm{one anonymous reviewer}}
\end{review}
%
%\dedicated{}
%
% HISTORY:
\received{\sday{13} \smonth{8} \syear{2021}}
\accepted{\sday{23} \smonth{9} \syear{2021}}
%
\begin{abstract}
%\parttitle{}
An increasing number of researchers support reproducibility by
including pointers to and
descriptions of datasets, software and methods in their publications.
However, scientific
articles may be ambiguous, incomplete and difficult to process by
automated systems. In
this paper we introduce RO-Crate, an open, community-driven, and
lightweight approach
to packaging research artefacts along with their metadata in a machine
readable manner.
RO-Crate is based on \surl{Schema.org} annotations in JSON-LD, aiming to
establish best
practices to formally describe metadata in an accessible and practical
way for their
use in a wide variety of situations.

An RO-Crate is a structured archive of all the items that contributed
to a research
outcome, including their identifiers, provenance, relations and
annotations. As a general
purpose packaging approach for data and their metadata, RO-Crate is
used across multiple
areas, including bioinformatics, digital humanities and regulatory
sciences. By applying
``just enough'' Linked Data standards, RO-Crate simplifies the process of
making research outputs FAIR while also enhancing research reproducibility.

An RO-Crate for this article\footnote{\url
{https://w3id.org/ro/doi/10.5281/zenodo.5146227}}
is archived at \url{https://doi.org/10.5281/zenodo.5146227}.
\end{abstract}
%
\begin{keyword} %% 1 u.c.
\kwd{Data publishing}
\kwd{data packaging}
\kwd{FAIR}
\kwd{Linked Data}
\kwd{metadata}
\kwd{reproducibility}
\kwd{research object}
\end{keyword}
%
\end{frontmatter}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Main text entry area:
% Every *.md file in content/ in order, except 00, 90
\section{Introduction}

\label{sec:introduction}

The move towards Open Science has increased the need and demand for the
publication of artefacts of the research process
\cite{sefton_blog_post_2021}. This is particularly apparent in domains that
rely on computational experiments; for example, the publication of
software, datasets and records of the dependencies that such
experiments rely on \cite{doi:10.1126/science.aah6168}.

It is often argued that the publication of these assets, and
specifically software
\cite{doi:10.3233/DS-190026}, workflows \cite{doi:10.1162/dint_a_00033} and
data, should follow the FAIR principles \cite{doi:10.1038/sdata.2016.18};
namely, that they are Findable, Accessible, Interoperable and Reusable.
These principles are agnostic to the \textit{implementation} strategy needed
to comply with them. Hence, there has been an increasing amount of work
in the development of platforms and specifications that aim to fulfil
these goals \cite{isbn:9781315351148}.

Important examples include data publication with rich metadata (e.g.
Zenodo \cite{doi:10.3897/biss.3.37080}), domain-specific data deposition
(e.g. PDB \cite{doi:10.1093/nar/gkl971}) and following practices for
reproducible research software \cite{doi:10.1371/journal.pcbi.1003285}
(e.g. use of containers). While these platforms are useful, experience
has shown that it is important to put greater emphasis on the
interconnection of the multiple artefacts that make up the research
process \cite{doi:10.1016/j.ijhcs.2020.102562}.

The notion of \textbf{Research Objects} \cite{doi:10.1016/j.future.2011.08.004}
(RO) was introduced to address this connectivity, providing
semantically rich \textit{aggregations} of (potentially distributed) resources
with a layer of structure over a research study; this is then to be
delivered in a \textit{machine-readable format}.

A Research Object combines the ability to bundle multiple types of
artefacts together, such as spreadsheets, code, examples, and figures.
The RO is augmented with annotations and relationships that describe
the artefacts' \textit{context} (e.g. a CSV being used by a script, a figure
being a result of a workflow).

This notion of ROs provides a compelling vision as an approach for
implementing FAIR data. However, existing Research Object
implementations require a large technology stack
\cite{doi:10.1016/j.websem.2015.01.003}, are typically tailored to a
particular platform and are also not easily usable by end-users.

To address this gap, a new community came together
\cite{doi:10.5281/zenodo.3250687} to develop \textbf{RO-Crate} -- an \textit{approach
to package and aggregate research artefacts with their metadata and
relationships}. The aim of this paper is to introduce RO-Crate and
assess it as a strategy for making multiple types of research artefacts
FAIR. Specifically, the contributions of this paper are as follows:
%
\begin{enumerate}
\item[1.] An introduction to RO-Crate, its purpose and context;
\item[2.] A guide to the RO-Crate community and tooling;
\item[3.] Examples of RO-Crate usage, demonstrating its value as connective
tissue for different artefacts from different communities.
\end{enumerate}

The rest of this paper is organised as follows. We first describe
RO-Crate through its development methodology that formed the RO-Crate
concept, showing its foundations in Linked Data and emerging
principles. We then define RO-Crate technically, before we introduce
the community and tooling. We move to analyse RO-Crate with respect to
usage in a diverse set of domains. Finally, we present related work and
conclude with some remarks including RO-Crate highlights and future
work. The appendix adds a formal definition of RO-Crate using
First-Order logic.

\section{RO-Crate} %{#rocrate}

\label{sec:rocrate}

RO-Crate aims to provide an approach to packaging research artefacts
with their metadata that can be easily adopted. To illustrate this, let
us imagine a research paper reporting on the sequence analysis of
proteins obtained from an experiment on mice. The sequence output
files, sequence analysis code, resulting data and reports summarising
statistical measures are all important and inter-related research
artefacts, and consequently would ideally all be co-located in a
directory and accompanied with their corresponding metadata. In
reality, some of the artefacts (e.g. data or software) will be recorded
as external reference to repositories that are not necessarily
following the FAIR principles. This conceptual directory, along with
the relationships between its constituent digital artefacts, is what
the RO-Crate model aims to represent, linking together all the elements
of an experiment that are required for the experiment's reproducibility
and reusability.

The question then arises as to how the directory with all this material
should be packaged in a manner that is accessible and usable by others.
This means programmatically and automatically accessible by machines
and human readable. A de facto approach to sharing collections of
resources is through compressed archives (e.g. a zip file). This solves
the problem of ``packaging'', but it does not guarantee downstream
access to all artefacts in a programmatic fashion, nor describe the
role of each file in that particular research. Both features, the
ability to automatically access and reason about an object, are crucial
and lead to the need for explicit metadata about the contents of the
folder, describing each and linking them together.

Examples of metadata descriptions across a wide range of
domains\footnote{\url{https://rdamsc.bath.ac.uk/scheme-index}} abound within the
literature, both in research data management
\cite{doi:10.1007/s10209-016-0475-y,farnel_2014,doi:10.2777/620649}
and within library and information
systems\footnote{\url{https://www.loc.gov/librarians/standards}} \cite{chan_1995,doi:10.1515/9783598441844}. However, many of these approaches require
knowledge of metadata schemas, particular annotation systems, or the
use of complex software stacks. Indeed, particularly within research,
these requirements have led to a lack of adoption and growing
frustration with current tooling and specifications
\cite{neylon_blog_post_2017,doi:10.1007/s00267-014-0258-2,doi:10.1038/s41597-020-0524-5}.

RO-Crate seeks to address this complexity by:
%
\begin{enumerate}
\item[1.] being conceptually simple and easy to understand for developers;
\item[2.] providing strong, easy tooling for integration into community projects;
\item[3.] providing a strong and opinionated guide regarding current best practices;
\item[4.] adopting de-facto standards that are widely used on the Web.
\end{enumerate}

In the following sections we demonstrate how the RO-Crate specification
and ecosystem achieve these goals.

%% sub-subsections below to cater for figures in
%% markdown vs LaTex
\subsection{Development methodology}

\label{sec:methodology}

It is a good question as to what base level we assume for `conceptually
simple'. We take simplicity to apply at two levels: for the
\textit{developers} who produce the platforms and for the \textit{data practitioners}
and users of those platforms.

For our development methodology we followed the mantra of working
closely with a small group to really get a deep understanding of
requirements and ensure rapid feedback loops. We created a pool of
early adopter projects from a range of disciplines and groups,
primarily addressing developers of platforms. Thus the base level for
simplicity was \textbf{developer friendliness}.

We assumed a developer familiar with making Web applications with JSON
data (who would then learn how to make \textit{RO-Crate JSON-LD}), which
informed core design choices for our JSON-level documentation approach
and RO-Crate serialization (Section~\ref{sec:implementation}). Our
group of early adopters, growing as the community evolved, drove the
RO-Crate requirements and provided feedback through our multiple
communication channels including bi-monthly meetings, which we describe
in Section~\ref{sec:community} along with the established norms.

Addressing the simplicity of understanding and engaging with RO-Crate
by data practitioners is through the platforms, for example with
interactive tools (Section~\ref{sec:tooling}) like
Describo\footnote{\url{https://uts-eresearch.github.io/describo/}} \cite{describo} and
Jupyter notebooks \cite{doi:10.3233/978-1-61499-649-1-87}, and by close
discussions with domain scientists on how to appropriately capture what
they determine to be relevant metadata. This ultimately requires a new
type of awareness and learning material separate from developer
specifications, focusing on the simplicity of extensibility to serve
the user needs, along with user-driven development of new RO-Crate
Profiles specific for their needs (Section~\ref{sec:inuse}).

\subsection{Conceptual definition}% {#conceptual}

\label{sec:conceptual}

A key premise of RO-Crate is the existence of a wide variety of
resources on the Web that can help describe research. As such, RO-Crate
relies on the Linked Data principles
\cite{doi:10.2200/S00334ED1V01Y201102WBE001}. Figure~\ref{fig:conceptual}
shows the main conceptual elements involved in an RO-Crate: The
RO-Crate Metadata File (top) describes the Research Object using
structured metadata including external references, coupled with the
contained artefacts (bottom) bundled and described by the RO-Crate.

The conceptual notion of a \textit{Research Object}
\cite{doi:10.1016/j.future.2011.08.004} is thus realized with the RO-Crate
model and serialized using Linked Data constructs within the RO-Crate
metadata file.

\begin{figure}%[t!]
\includegraphics{210053f01}
\caption{Conceptual overview of RO-Crate. A \emph{Persistent
Identifier} (PID) \cite{doi:10.1371/journal.pbio.2001414} points to a
\emph{Research Object} (RO), which may be archived using different
packaging approaches like BagIt \cite{doi:10.17487/rfc8493}, OCFL \cite
{ocfl_2020}, git or ZIP. The RO is described within a \emph{RO-Crate
Metadata File}, providing identifiers for \emph{authors} using ORCID,
\emph{organisations} using Research Organization Registry (ROR) \cite
{doi:10.6087/kcse.192} and licences such as Creative Commons using SPDX
identifiers. The \emph{RO-Crate content} is further described with
additional metadata following a Linked Data approach. Data can be
embedded files and directories, as well as links to external Web
resources, PIDs and nested RO-Crates.}
\label{fig:conceptual}
\end{figure}

\subsubsection{Linked Data as a foundation}% {#linkeddata}

\label{sec:linkeddata}

The \textbf{Linked Data} principles \cite{doi:10.4018/978-1-60960-593-3.ch008}
(use of IRIs\footnote{\textbf{IRI}s \cite{doi:10.17487/rfc3987} are a generalisation
 of \textit{URI}s
(which include well-known http/https URLs), permitting international
Unicode characters without percent encoding, commonly used on the
browser address bar and in HTML5.} to identify resources (i.e. artefacts), resolvable via
HTTP, enriched with metadata and linked to each other) are core to
RO-Crate; therefore IRIs are used to identify an RO-Crate, its
constituent parts and metadata descriptions, and the properties and
classes used in the metadata.

RO-Crates are \textit{self-described} and follow the Linked Data principles to
describe all of their resources in both human and machine readable
manner. Hence, resources are identified using global identifiers
(absolute IRIs) where possible; and relationships between two resources
are defined with links.

The foundation of Linked Data and shared vocabularies also means that
multiple RO-Crates and other Linked Data resources can be indexed,
combined, queried, validated or transformed using existing Semantic Web
technologies such as SPARQL,\footnote{\url{https://www.w3.org/TR/sparql11-overview}}
SHACL\footnote{\url{https://www.w3.org/TR/shacl/}} and well established \textit{knowledge
graph} triple stores like Apache Jena\footnote{\url{https://jena.apache.org/}} and
OntoText GraphDB.\footnote{\url{https://www.ontotext.com/products/graphdb/}}

The possibilities of consuming\footnote{Some consideration is needed in processing of RO-Crates as
knowledge graphs, e.g. establishing absolute URIs for files inside a
ZIP archive, detailed in the RO-Crate specification: \url
{https://www.researchobject.org/ro-crate/1.1/appendix/relative-uris.html}.} RO-Crate metadata with such
powerful tools gives another strong reason for using Linked Data as a
foundation. This use of mature Web\footnote{Note that an RO-Crate is not required to be published on the
Web, see Section~\ref{sec:selfdescribed}.} technologies also means its
developers and consumers are not restricted to the Research Object
aspects that have already been specified by the RO-Crate community, but
can extend and integrate RO-Crate in multiple standardized ways.

\subsubsection{RO-Crate is a self-described container}% {#selfdescribed}

\label{sec:selfdescribed}

An RO-Crate is defined\footnote{\url{https://www.researchobject.org/ro-crate/1.1/structure.html\#ro-crate-metadata-file-ro-crate-metadatajson}}
as a self-described \textbf{Root Data Entity} that describes and contains
\textit{data entities}, which are further described by referencing {contextual
entities}. A \textbf{data entity} is either a \textit{file} (i.e. a byte sequence
stored on disk somewhere) or a \textit{directory} (i.e. set of named files and
other directories). A~file does not need to be stored inside the
RO-Crate root, it can be referenced via a PID/IRI. A \textbf{contextual
entity} exists outside the information system (e.g. a Person, a
workflow language) and is stored solely by its metadata. The
representation of a \textbf{data entity} as a byte sequence makes it
possible to store a variety of research artefacts including not only
data but also, for instance, software and text.

The Root Data Entity is a directory, the \textit{RO-Crate Root}, identified by
the presence of the \textbf{RO-Crate Metadata File} \texttt{ro-crate-metadata.json}
(top of Fig.~\ref{fig:conceptual}). This is a JSON-LD file that
describes the RO-Crate, its content and related metadata using Linked
Data. JSON-LD is a W3C standard RDF serialisation that has become
popular as it is easy to read by humans while also offers some
advantages for data exchange on the Internet. JSON-LD, a subset of the
widely supported and well-known JSON format, has tooling available for
many programming languages.\footnote{\url{https://json-ld.org/\#developers}}

The minimal requirements for the root data entity
metadata\footnote{\url{https://www.researchobject.org/ro-crate/1.1/root-data-entity.html\#direct-properties-of-the-root-data-entity}}
are \texttt{name}, \texttt{description} and \texttt{datePublished}, as well as a contextual
entity identifying its \texttt{license} -- additional metadata are commonly
added to entities depending on the purpose of the particular RO-Crate.

RO-Crates can be stored, transferred or published in multiple ways,
e.g. BagIt \cite{doi:10.17487/rfc8493}, Oxford Common File Layout
\cite{ocfl_2020} (OCFL), downloadable ZIP archives in Zenodo or through
dedicated online repositories, as well as published directly on the
Web, e.g. using GitHub Pages.\footnote{\url{https://pages.github.com/}} Combined
with Linked Data identifiers, this caters for a diverse set of storage
and access requirements across different scientific domains, from
metagenomics workflows producing hundreds of gigabytes of genome data
to cultural heritage records with access restrictions for personally
identifiable data. Specific \textit{RO-Crate profiles} (Section~\ref{sec:profiles}) may constrain serialization and publication
expectations, and require additional contextual types and properties.

\subsubsection{Data entities are described using contextual entities}% {#contextualentities}

\label{sec:contextualentities}

RO-Crate distinguishes between data and contextual
entities\footnote{\url{https://www.researchobject.org/ro-crate/1.1/contextual-entities.html\#contextual-vs-data-entities}}
in a similar way to HTTP terminology's early attempt to separate
\textit{information} (data) and \textit{non-information} (contextual) resources
\cite{httprange14}. Data entities are usually files and directories located
by relative IRI references within the RO-Crate Root, but they can also
be Web resources or restricted data identified with absolute IRIs,
including \textit{Persistent Identifiers} (PIDs) \cite{doi:10.1371/journal.pbio.2001414}.

As both types of entities are identified by IRIs, their distinction is
allowed to be blurry; data entities can be located anywhere and be
complex, while contextual entities can have a Web presence beyond their
description inside the RO-Crate. For instance
\texttt{https://orcid.org/0000-0002-1825-0097} is primarily an identifier for
a person, but secondarily it is also a Web page and a way to refer to
their academic work.

A particular IRI may appear as a contextual entity in one RO-Crate and
as a data entity in another; the distinction lies in the fact that data
entities can be considered to be \textit{contained} or captured by that
RO-Crate (\textit{RO Content} in Fig.~\ref{fig:conceptual}), while
contextual entities mainly \textit{explain} an RO-Crate or its content
(although this distinction is not a formal requirement).
